A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-10-21 evaluation.tex
Text/tightening

The benchmarks are executed in parallel across a cluster of identical
machines, \BenchmarkRig{}, but as the sole job for each machine.  We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \numprint{10}s on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout.

@@

In \cref{fig:solve-division}, we show how many of the~\NrBenchmarks{} instances
the respective back-end could solve. A summary of their outcomes by instance
type is also available in \cref{tab:solve-status}. Note that many instances lack
a corum ground truth as they are solved by only one backend. We see that
\Calculus{} generally outperforms \Nuxmv{} on determining unsatisfiability, as
does baseline, while being similar at satisfiable instances. Both \Calculus{} and
\Nuxmv{} outperforms baseline on satisfiable instances. On satisfiable
instances, \Nuxmv{} and \Calculus{} have similar performance.

Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early at a heavy penalty to satisfiable
instances roughly similar to \cite{approximate-parikh}. The heuristic is enabled
since the improvement is significant compared to the extra cost and since we
have observed more unsatisfiable instances than satisfiable ones.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, PC* could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than PC*.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the PC* backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for PC*. On the other hand, PC* has a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \numprint{120}s-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, PC* and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
  \caption{Evaluation of \Calculus{} solving Parikh automata problems generated by \Ostrich{}.}
\end{figure}

\subsubsection{Scalability}\label{sec:scaling}%
To investigate scaling we additionally execute \numprint{2000} randomly sampled
(without replacement) benchmarks with a \numprint{120}-second timeout. In
addition to the previous back-ends, we also execute \Calculus{} with clause
learning and restarts disabled, and with only clause learning disabled to show
the impact of \cref{sec:random-restarts}, and \cref{sec:clause-learning}
respectively.

A cactus plot showing the number of instances solved within a given timeout can
be seen in \cref{fig:cactus}. We see that \Calculus{} outperforms \Nuxmv{} in
general but that \Nuxmv{} might scale better on very long runtimes.
Non-systematic observations suggest that \Nuxmv{}'s scaling benefits kicks in
faster on newer machines, so it is likely that the amount of CPU cycles spent is
the underlying factor. This is likely due to two factors. First, \Nuxmv{} is
more mature than \Calculus{}, and it makes sense that its more general model
checking methods would pay off on more difficult instances where more
problem-specific methods offer less payoff. Second, for longer running
calculations clause learning as described in \cref{sec:clause-learning} can be
assumed to matter more. However, the clauses we learn are still unpolished, and
in particular, generalise poorly across product computations. Improvements are
left as future work.

@@

\begin{table}
  \begin{center}
  \input{experiments/qf_slia_comparison.tex}
  \end{center}
  \caption{Number of solved benchmarks from the single-query QF\_SLIA
    track of quantifier-free string linear integer arithmetic at
    SMT-COMP 2023. The numbers for each solver are from the publicly
    available data files, except for CA-Str, which is executed by us
    (600s timeout, AMD Opteron 2220
    SE), and \Ostrich+CA, which is a virtual portfolio. }
  \label{tab:solve-status-smt-comp}
\end{table}

To evaluate the effectiveness of \Calculus{} in an actual string solver,
\Catra{} was experimentally integrated as the Parikh automata product solver
into the string solver \Ostrich{}~version~1.3. \Ostrich{} is an independently
developed solver that participated in the recent SMT-COMP~2023\footnote{\url{https://smt-comp.github.io/2023/participants/ostrich}},
 winning the single-query track
for quantifier-free strings (\texttt{QF\_S}), as well as dominating
other solvers on the complete set of
 unsatisfiable string benchmarks~\cite{smt-comp-23}.

 At SMT-COMP, \Ostrich{} competed using a portfolio of three different
 back-ends:
\begin{itemize}
\item \textbf{BW-Str:} a backward-propagation-based solver, not utilising
  Parikh automata~\cite{ostrich}.
\item \textbf{ADT-Str:} a solver based on algebraic data-types.
\item \textbf{CE-Str:} a re-implementation of the \OstrichPlus{}
  algorithm~\cite{ostrich-plus}, using Parikh automata and the
  encoding from~\cite{generate-parikh-image}.
\end{itemize}

@@

The results show that \Calculus{} can be used to enhance the
performance of an automaton-based string solver.
The results should be considered as preliminary, however, 
as we believe that a deeper integration of our \Catra{} solver into
string solvers can lead to significant performance gains.

@@

The most obvious threats to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Since addressing that bug we have observed no other soundness
issues, despite re-executing the same instances multiple times for long
durations.

>>>>

The benchmarks are executed in parallel across a cluster of identical
machines, \BenchmarkRig{}, as the sole job for each machine.  We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \numprint{10}s on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout.

@@

In \cref{fig:solve-division}, we show how many of the~\NrBenchmarks{} instances
the respective back-end could solve. A summary of their outcomes by instance
type is also available in \cref{tab:solve-status}. Note that many instances lack
a ground truth as they are solved by only one backend. We see that
\Calculus{} generally outperforms \Nuxmv{} on determining unsatisfiability, as
does baseline, while being similar at satisfiable instances. Both \Calculus{} and
\Nuxmv{} outperforms baseline on satisfiable instances. On satisfiable
instances, \Nuxmv{} and \Calculus{} have similar performance.

Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early, similar to~\cite{approximate-parikh}.
The heuristic is enabled since the improvement is significant compared to the
extra cost.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, PC* could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than PC*.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Total numbers of solved instances differ.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the PC* backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for PC*. On the other hand, PC* has a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \numprint{120}s-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, PC* and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
  \caption{Evaluation of \Calculus{} solving Parikh automata problems generated by \Ostrich{}.}
\end{figure}

\subsubsection{Scalability}\label{sec:scaling}%
To investigate scaling we additionally execute \numprint{2000} randomly sampled
(without replacement) benchmarks with a \numprint{120}-second timeout. We add
runs of \Calculus{} with clause learning and restarts disabled, and with only
clause learning disabled to show the impact of \cref{sec:random-restarts}, and
\cref{sec:clause-learning} respectively.

A cactus plot showing the number of instances solved within a given timeout can
be seen in \cref{fig:cactus}. We see that \Calculus{} outperforms \Nuxmv{} in
general but that \Nuxmv{} might scale better on very long runtimes. This is
likely due to two factors. First, \Nuxmv{} is more mature than \Calculus{}, and
its more general model checking methods might pay off on more difficult
instances compared to problem-specific methods. Second, for longer-running
calculations clause learning as described in \cref{sec:clause-learning} might
matter more. As clause learning in \Catra{} is rudimentary and generalises
poorly across product computations, there is a performance hit.

@@

\begin{table}
  \begin{center}
  \input{experiments/qf_slia_comparison.tex}
  \end{center}
  \caption{Number of solved benchmarks in the set of quantifier-free string
    linear integer arithmetic (QF\_SLIA) at SMT-COMP 2023. The numbers are from
    the competition results, except for CA-Str, which is executed by us (600s
    timeout, AMD Opteron 2220 SE), and \Ostrich+CA, a virtual portfolio. }
  \label{tab:solve-status-smt-comp}
\end{table}

To evaluate the effectiveness of \Calculus{} in a string solver, \Catra{} was
experimentally integrated as the Parikh automata product solver into the string
solver \Ostrich{}~version~1.3. \Ostrich{} is an independently developed solver
that participated in the recent
SMT-COMP~2023\footnote{\url{https://smt-comp.github.io/2023/participants/ostrich}},
winning the single-query track for quantifier-free strings (\texttt{QF\_S}), as
well as dominating other solvers on unsatisfiable string
benchmarks~\cite{smt-comp-23}.

 At SMT-COMP, \Ostrich{} competed using a portfolio of three different
 back-ends:
\begin{itemize}
\item \textbf{BW-Str:} a backward-propagation-based solver, not utilising
  Parikh automata~\cite{ostrich}.
\item \textbf{ADT-Str:} a solver based on algebraic data-types.
\item \textbf{CE-Str:} a re-implementation of the \OstrichPlus{}
  algorithm~\cite{ostrich-plus}, using Parikh automata and the
  encoding from~\cite{generate-parikh-image} (referred to in this paper as baseline).
\end{itemize}

@@

The results show that \Calculus{} can be used to enhance the performance of an
automaton-based string solver. The results should be considered preliminary,
however, as we believe that a deeper integration of \Catra{} into string solvers
can lead to significant performance gains.

@@

The most obvious threat to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Since addressing that bug we have observed no further soundness
issues.
